---
title: "Assignment 1"
output: html_notebook
---

```{r}
df <- read.csv(file="Assignment1_Data.csv")
```

### Question 1
For each of the 17 variables in the data file, indicate of what type they are and
explain why. Choose from nominal scale, ordinal scale, interval scale or ratio scale.
```{r}
data_types <- function(frame) {
  res <- lapply(frame, class)
  res_frame <- data.frame(unlist(res))
  barplot(table(res_frame), main="Data Types", col="steelblue", ylab="Number of Features")
}
data_types(df)
```
### Question 2
What is the percentage of companies that have introduced new or significantly
improved products/goods?
```{r}
ProdInn1 <- df[(df$ProductInnovation==1),]
ProdInn0 <- df[(df$ProductInnovation==0),]

perc <- dim(ProdInn1)[1]/dim(df)[1]*100
cat(round(perc,2), "%")
```

### Question 3
Compare the distribution of R&D expenditures for companies that have introduced
new or significantly improved products/goods, versus companies that have not.
Use a plot and a statistical test. Explain which plot and test you use, and why.
What do you conclude?
```{r}
# heavily skewed to the right
summary(ProdInn1$RDExp)
summary(ProdInn0$RDExp)
```

```{r}
# Take the log of the data so the distribution can be visualized better

hist(log(ProdInn1$RDExp), breaks=100, col='blue')
hist(log(ProdInn0$RDExp), breaks=100, col='red')
```
```{r}

# The RDExp of companies who did innovate has a different distribution from that of those who did not innovate
ks.test(log(ProdInn1$RDExp), log(ProdInn0$RDExp))
```


### Question 4
Is manufacturing innovation (i.e., new or significantly improved manufacturing
process and/or procedures for service delivery) more common for companies that
have introduced new or significantly improved products/goods, versus companies
that have not? Support your answer.

```{r}
# Manufacturing innovation is not more common for companies that innovated their products/goods, versus companies that did not
cat(dim(ProdInn1[ProdInn1$ManufacturingInnovation==1,])[1]/dim(ProdInn1)[1]*100,"% product=1, manuf=1\n")
cat(dim(ProdInn1[ProdInn1$ManufacturingInnovation==0,])[1]/dim(ProdInn1)[1]*100,"% product=1, manuf=0\n")

cat(dim(ProdInn0[ProdInn1$ManufacturingInnovation==1,])[1]/dim(ProdInn0)[1]*100,"% product=0, manuf=1\n")
cat(dim(ProdInn0[ProdInn1$ManufacturingInnovation==0,])[1]/dim(ProdInn0)[1]*100,"% product=0, manuf=0\n")
```

### Question 5
Which variable(s) contain missing values?
```{r}
# NrEmployees has missing values
colSums(is.na(df))
```

### Question 6
What is the percentage of missingness?
```{r}
cat(mean(is.na(df))*100,"% of all data points are missing.\n")
cat(mean(is.na(df$NrEmployees))*100, "% of the NrEmployees column is missing.")
```

### Question 7
Is the missingness related to any of the other variables in the data set? Explain
which tests you use and why.
```{r}
# Missingness is not related to any of the other variables in the dataset.
# Used 2-sample t-test to compare the mean of each variable of complete cases with that of incomplete cases

missing <- df[!complete.cases(df),]
complete <- df[complete.cases(df),]

for (i in colnames(df)){
  if(i!="NrEmployees"){
    print(i)
    print(t.test(missing[[i]],complete[[i]])$p.value)
  }
}
```

### Question 8
What do you conclude about the type of missingness from your tests of the previous
question?

I conclude that the data is missing at random and can be omitted from the dataset.

### Question 9
Compute the Mahalanobis distances of the companies based on the information
variables. Graphically present the Mahalanobis distances.

```{r}
maha_distance <- complete
dist <- mahalanobis(complete, center=colMeans(complete), cov=cov(complete))
maha_distance$dist <- round(dist, 2)

maha_distance$outlier_maha <- 0
maha_distance$outlier_maha[maha_distance$dist >3] <- 1

upper.whisker <- boxplot(maha_distance$dist)$stats[5]
```

### Question 10
Can we use the rule-of-thumb that observations of which the Mahalanobis Distance
divided q is larger 3 are outliers? Here, q is the number of variables used in the
calculation of the Mahalanobis distance. Explain why or why not.

```{r}
# Do not use 3 as the rule of thumb because the data is not normally distributed.
mean(maha_distance$outlier_maha)
```


### Question 11
Which companies, if any, are outliers? Describe the decision process by which you
have or have not identified outliers. If you have identified outliers, perform the
rest of the assignment after dropping the outliers from the analysis.

```{r}
# only 6 outliers when using the upper whisker as the threshold (54.57)
outliers <- maha_distance[maha_distance$dist>upper.whisker,c(1:17)]
no_outliers <- maha_distance[maha_distance$dist<=upper.whisker,c(1:17)]
```

### Question 12
Is a factor analysis appropriate for the questions on the information sources of
innovations? Explain how you have checked whether factor analysis is appropriate.

```{r}
library(psych)
library(ggplot2)
# make heatmap
cormat <- round(cor(no_outliers),2)

melted_cormat <- reshape2::melt(cormat)
head(melted_cormat)

ggplot(data = melted_cormat, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile()
```


### Question 13
Make a scree plot. How many factors does the scree plot suggest?
```{r}
# 2 factors
EigenDecomp <- eigen(cov(no_outliers))
EigenValues <- EigenDecomp$values
plot(EigenValues,main="Scree Plot IAS",type="l", xlab="Number of Factors",ylab="Eigen Values")
```


### Question 14
For what purpose is a rotation of the factors in an exploratory factor analysis used?
Should we perform a rotation in this setting? Why or why not?

Rotate so that the factor loadings are easier to interpret. For example, consider 2 factor loadings F1 and F2 with variables X1 and X2 in a cluster with higher loadings for F1 and X3 and X4 with higher loadings for F2. When we perform a varimax rotation, we will create new factor loadings that give variables X1 and X2 high values for F1 and low for F2, and vice versa for variables X3 and X4.

### Question 15
Perform factor analysis using the number of factors as suggested by the scree plot.
Clearly state how many factors you have included and whether or not you have
used a rotation. Report the estimated uniquenesses and factor loadings.

```{r}
# Two factors are not sufficent. Chi squared null hypotheses is rejected.
k<-2
EFA.kfactors <- factanal(no_outliers, factors=k, rotation="none")
EFA.kfactors$converged
EFA.kfactors
```

### Question 16
Check whether all the variables should be retained in the factor analysis.
(a) Which criteria do you use? Which of the variables, if any, should be dropped?

```{r}
# Drop variables with uniquness > 0.8
# Consider lowering to .72 so that productinnovation is also dropped
# I think we should drop product innovation because it is otherwise grouped into factor with the Info variables, which does not make much sense as the survey questions are on different topics.
# Rerun analysis without these variables
EFA.kfactors <- factanal(no_outliers, factors=2, rotation="none")
EFA.kfactors$uniquenesses[EFA.kfactors$uniquenesses>.8]
```

(b) If one or more of the variables should be dropped, redo the scree plot and the
estimation of the factor model with the number of factors suggested by the
scree plot.


```{r}
# drop 5 variables about Innovationand redo Scree plot
# plot indicates 2 Factors
EigenDecomp <- eigen(cov(no_outliers[,c(1:2,8:17)]))
EigenValues <- EigenDecomp$values
plot(EigenValues,main="Scree Plot IAS",type="l", xlab="Number of Factors",ylab="Eigen Values")
```

```{r}
k<-2
EFA.kfactors <- factanal(no_outliers[,c(1:2,8:17)], factors=k, rotation="varimax")
EFA.kfactors$converged
# reject null hypothesis that 2 factors is sufficent. 
EFA.kfactors
```

(c) If needed, make further changes to the factor model based on the output of
the factor analysis. Report and motivate all the steps you have taken to arrive
at a final factor model. How many factors are included in your final model?

I would choose 4 factors in the final model, despite the failed chi-squared test because 4 factors gives communality for all factor loadings and only variable Info5 noew not have a high loading on one factor.

```{r}
# The Chi-squared test suggests 7 factors, but we need to make sure that makes sense when looking at the factor loadings.
NrFactors <- 3:7
pvals <- rep(NA,times=length(NrFactors))
for(k in NrFactors){
  # Estimate k-factor model:
  Estimation <- factanal(no_outliers[,c(1:2,8:17)],factors=k,rotation="none")
  pvals[k] <- Estimation$PVAL
}

cbind(NrFactors,pvals)
```

Each factor should have a high loading on several variables (communality) and each variable should load high on one factor loading and variable groupings should make sense.

```{r}

# Factors 4-7 don't have high communality
print("k=7")
EFA.kfactors <- factanal(no_outliers[,c(1:2,8:17)], factors=7, rotation="varimax")
EFA.kfactors$loadings

# Factors 4-6 don't have high communality
print("k=6")
EFA.kfactors <- factanal(no_outliers[,c(1:2,8:17)], factors=6, rotation="varimax")
EFA.kfactors$loadings

# Factor 5 doesn't have high communality
# Info 5, 8, and 10 do not score very high on one factor
print("k=5")
EFA.kfactors <- factanal(no_outliers[,c(1:2,8:17)], factors=5, rotation="varimax")
EFA.kfactors$loadings

# Factor 4 doesn't have high communality
# Info 5 and 9 do not score high on one factor
# Explained variance is getting low
print("k=4")
EFA.kfactors <- factanal(no_outliers[,c(1:2,8:17)], factors=4, rotation="varimax")
EFA.kfactors$loadings

# All factors have high communality
# All variables score high on one factor
# Explained variance is low
print("k=3")
EFA.kfactors <- factanal(no_outliers[,c(1:2,8:17)], factors=3, rotation="varimax")
EFA.kfactors$loadings
```

(d) Report the set of included variables, the estimated uniquenesses and factor
loadings of your final factor model. Continue the next questions with the
model you prefer.

```{r}
EFA.kfactors <- factanal(no_outliers[,c(1:2,8:17)], factors=5, rotation="varimax")
EFA.kfactors
```


### Question 17
Which names do you suggest for the factors? Motivate the names you have chosen.

5 Factors
1. Info1, Info 3, Info4, Info5
  - Sources within your firm or enterprise group
  - Customers or principals
  - Competitor or other enterprise in your industry sector
  - Consulting firms, private R&D business, commercial laboratories
2.  Info8, Info9, Info10
  - Conferences, fairs, exhibitions
  - Scientific periodicals and specialty publications
  - Associations and boards
3. Info6, Info7
  - Universities, universities of applied science or other higher education institutions
  - National or non-profit research institutions
4. RDExp, NrEmployees
5. Info2
  - Suppliers of facilities, equipment, components, software and services

3 Factors
1. Info1, Info2, Info 3, Info4, Info5, Info8, Info9, Info10
  - Suppliers of facilities, equipment, components, software and services
  - Competitor or other enterprise in your industry sector
  - Conferences, fairs, exhibitions
  - Scientific periodicals and specialty publications
  - Associations and boards
  - Consulting firms, private R&D business, commercial laboratories
  - Sources within your firm or enterprise group
  - Customers or principals
2. Info6, Info7
  - Universities, universities of applied science or other higher education institutions
  - National or non-profit research institutions
3. RDExp, NrEmployees



### Question 18
Perform a reliability analysis of each factor.
(a) Which criterion do you use? Which of the variables, if any, should be
dropped?
```{r}
CalphaTimeEscalation <- alpha(no_outliers[,c(1:2,8:17)])
CalphaTimeEscalation
```

(b) If one or more of the variables should be dropped, redo the scree plot and
the estimation of the factor model with the number of factors suggested by
the scree plot. Report and motivate all the steps you have taken to arrive
at a final factor model. Report the set of included variables, the estimated
uniquenesses and factor loadings of your final factor model.

### Question 19
Perform a validity analysis of your factor model by splitting the sample in two ran-
dom partitions and comparing the factor loadings. Report the estimated unique-
nesses and factor loadings of the split samples. What do you conclude?

### Question 20
Report the results of the factor analysis in a table, including all the important
metrics.

### Question 21
What type of variables are the estimated factors? Choose from nominal scale,
ordinal scale, interval scale or ratio scale and explain your answer.

### Question 22
For each type of innovation (product, service, manufacturing, logistics and sup-
port), check whether firms that have introduced such innovations load higher on 
each of the factors than firms that have not. Support your answers by plots and
by statistical tests. What do you conclude?

### Question 23
Is there a strong correlation between the factor scores and R&D expenditures?
Support your answer by a plot and by a statistical test. What do you conclude?
